---
title: "A SEIR model"
author: "Christian Barz"
date: "creation date: 29-07-2020, update date: `r lubridate::today()`"
output:
  html_document:
    fig_caption: yes
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    toc_float:
      smooth_scroll: FALSE
    theme: readable
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.path="Figs/", message=FALSE, warning=FALSE,
                      echo=TRUE, results="hide", fig.width=11)
```

# Introduction

## goal

The goal is to implement a SEIR model. I do not have any epidemilogical and can not


## setup and reproducibility

```{r}
library(pacman)
p_load(
 tidyverse,  # data wrangling
  lubridate, # date time operations
  outbreaks, # data
  rstan      # use Stan in R
)
```


## data

## data

In our example we are examine an outbreak of influenza A (H1N1) in 1978 at a british school. The data is freely available in the R package `outbreaks` and consists of the daily number of students in bed spanning over a time interval of 14 days.

There were 763 male students and 512 of them became ill. It is reported that one infected boy started the epidemic, which spreads in the relativly closed community.

We load the data and inspect it:

```{r DataLoad}
df <- influenza_england_1978_school
knitr::kable(df)
```


```{r DataInspection}
df %>% ggplot() + 
  geom_point(mapping = aes(x = date, y = in_bed)) +
  labs(y = "number of students in bed")
```


# Theoretical description

We consider a [SEIR-model](https://de.wikipedia.org/wiki/SEIR-Modell) given by the following system of differential equations:
  
$$
\begin{aligned}
\frac{dS}{dt} = & -\frac{\beta \cdot S \cdot I}{N}\\
\frac{dE}{dt} = & \frac{\beta \cdot S \cdot I}{N} - \alpha \cdot E\\
\frac{dI}{dt} = & \alpha\cdot E -\gamma\cdot I\\
\frac{dR}{dt} = & \gamma\cdot I
\end{aligned}
$$

The disease has started wich one infected individual and hence gives rise to the following initial conditions:

$$
\begin{aligned}
S(0)  =&  N -1\\
E(0)  =&  0\\
I(0)  =&  1\\
R(0)  =&  0
\end{aligned}
$$


# coding

```{r}

# define model

SEIR_model <- "functions {
  real[] sir(real t, real[] y, real[] theta, 
             real[] x_r, int[] x_i) {

      real S = y[1];
      real E = y[2];
      real I = y[3];
      real R = y[4];
      real N = x_i[1];
      
      real beta = theta[1];
      real alpha = theta[2];
      real gamma = theta[3];
      
      real dS_dt = -beta * I * S / N;
      real dE_dt =  beta * I * S / N - alpha * E;
      real dI_dt =  alpha * E - gamma * I;
      real dR_dt =  gamma * I;
      
      return {dS_dt, dE_dt, dI_dt, dR_dt};
  }
}
data {
  int<lower=1> n_days;
  real y0[3];
  real t0;
  real ts[n_days];
  int N;
  int cases[n_days];
}
transformed data {
  real x_r[0];
  int x_i[1] = { N };
}
parameters {
  real<lower=0> gamma;
  real<lower=0> beta;
  real<lower=0> phi_inv;
}
transformed parameters{
  real y[n_days, 3];
  real phi = 1. / phi_inv;
  {
    real theta[2];
    theta[1] = beta;
    theta[2] = gamma;

    y = integrate_ode_rk45(sir, y0, t0, ts, theta, x_r, x_i);
  }
}
model {
  //priors
  beta ~ normal(2, 1);
  gamma ~ normal(0.4, 0.5);
  phi_inv ~ exponential(5);
  
  //sampling distribution
  //col(matrix x, int n) - The n-th column of matrix x. Here the number of infected people 
  cases ~ neg_binomial_2(col(to_matrix(y), 2), phi);
}
generated quantities {
  real R0 = beta / gamma;
  real recovery_time = 1 / gamma;
  real pred_cases[n_days];
  pred_cases = neg_binomial_2_rng(col(to_matrix(y), 2), phi);
}

"
```


```{r}

```

